Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-04 paper/paper.tex
missing reference highlighter
moved cplx review to introduction; restructured and merged the review of kingma2015

\newcommand{\important}[1]{\textbf{\color{red} #1}}
\newcommand{\attn}[2]{\textbf{\color{red} #2~\textsuperscript{\textit{[#1]}}}}
\newcommand{\verify}[1]{\attn{verify}{#1}}
\newcommand{\rewrite}[1]{\attn{rewrite}{#1}}
\newcommand{\todo}[1]{{\color{blue} [TODO]} \important{#1}}

\newcommand{\real}{\mathbb{R}}
\newcommand{\cplx}{\mathbb{C}}
\newcommand{\tr}[1]{\mathop{tr}{#1}}

@@

@@

@@

For a particular family of variational posterior approximation, variational inference
can be used as a regularization method \cite{kingma_variational_2015} and as model sparsification
technique \cite{molchanov_variational_2017}. In \cite{kingma_variational_2015} the Binary
dropout and DropConnect \cite{hinton_improving_2012,dropconnect}, and Gaussian dropout
\cite{srivastava_dropout_2014,wang_fast_2013} were reframed as variational inference methods
by treating the induced parameter distributions as a factorized posterior approximations
of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where each component $q_{\theta}(\omega_j)$ is defined via $\omega_j = \mu_j \xi_j$
with $\xi_j \sim p(\xi_j)$ iid for some $p(\xi)$. In Binary dropout with rate $p \in (0, 1)$
the mutliplicative noise $\xi$ is distributed like $
  \mathcal{B}\bigl(
      \{0, \tfrac1{1-p}\}, 1-p
    \bigr)
$ whereas in Gaussian dropout $\mathcal{N}(1, \alpha)$ for $\alpha = \tfrac{p}{1-p}$.
Focusing on Gaussian dropout, \cite{kingma_variational_2015} proposed Variational Dropout
technique for Bayesian model regularization: $\alpha$ is treated as a variational parameter,
that enabled its optimization through SGVB. The variational approximation is a fully
factorized Gaussian distribution with components $
  q_{\theta}(\omega_j)
    = \mathcal{N}(\mu_j, \alpha_j \mu_j^2)
$. The flexibility in specifying the dependence of $q_\theta$ on $\alpha$ allows learning
structured dropout rates: global per network, local per layer and even individual, i.e.
specific to each parameter. In the case of a prior with improper log-uniform factors $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$, the KL-divergence term in \eqref{eq:elbo} as computed up to a constant term as a sum of
\begin{equation} \label{eq:improper-kl-div-real}
  KL(q_\theta(\omega) \| \pi(\omega))
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}

@@

The more recent developments of Bayesian dropout technique with factorized Gaussian
approximations are considered in \cite{molchanov_variational_2017,kharitonov_variational_2018}.
\cite{molchanov_variational_2017}
studies variational dropout from a model sparsification perspective, improve on the results
of \cite{kingma_variational_2015} in two ways: the authors propose a more accurate approximation
to the KL-divergence term, and propose additive noise parameterization for the factors of
$q_\theta(\omega)$, $\omega = W$. Under the original $(\mu, \alpha)$ parameterization the
stochastic output of a linear layer $
  y = W^\top x + b
$ with $
  W \in \mathbb{R}^{n\times m}
$ and independent $
  W_{ij} \sim \mathcal{N}(\mu_{ij}, \alpha_{ij} \mu_{ij}^2)
$ is equivalent in distribution to $
  y \sim \prod_i \mathcal{N}\Bigl(
      \sum_j \mu_{ij} x_j + b_i,
      \sum_j \alpha_{ij} \mu_{ij}^2 x_j^2
  \Bigr)
$. \cite{molchanov_variational_2017} argue that for the purposes of model sparsification this parameterization
should be abandoned in favour of an equivalent $(\mu, \sigma^2)$ with $\alpha$ calculated
using $\tfrac{\sigma^2}{\mu^2}$ when needed. This makes $\nabla_\mu \mathcal{L}^{SGVB}$,
which is particularly important for sparsification, independent of the noise injected by
the trick.

The KL-divergence term \eqref{eq:expect-improper-term-cplx} is half the log-moment of a non-central
$\chi^2$ distribution with shape $1$ and non-centrality $\lambda = \tfrac1{\alpha}$. In general,
there is no analytically tractable expression for log-moment of a $\chi^2_m$ with odd degrees of
freedom. To circumvent this, \cite{kingma_variational_2015} approximate the expectation by a
non-linear polynomial regression for truncated $\alpha \in (0, 1)$. This approximation was
significantly refined in \cite{molchanov_variational_2017}, by considering regression with sigmoid
and soft-plus basis functions fit to the Monte-Carlo estimator of \eqref{eq:improper-kl-div-real}
for $\alpha$ varying over a logarithmic grid. This enabled doing away with the clipping of the
dropout rate (relevance score) $\alpha$. Since gradient based methods rely on unbiased gradient
estimates, in sec.~\ref{sub:real-chisq-grad} we compare the derivative of the approximation
to the exact derivative of \eqref{eq:improper-kl-div-real} with respect to $\log \alpha$ and show
that the approximation in \cite{molchanov_variational_2017} is an adequate.

@@

The use $\cplx$-valued networks in applications is not new: earliest mentions include
\cite{citation_needed}. More recently they were applied to music transcription and speech
recognition \cite{trabelsi_deep_2017,1910.10202}. The key motivation behind adoption
of $\cplx$-networks is the fact that complex numbers are inherent in digital signal processing
\cite{citation_needed}, acoustic and radio signal analysis \cite{citation_needed}, and
quantum systems modelling \cite{citation_needed} etc. All these applications require
compliance with algebraic operations in field of complex numbers.

@@

Following \cite{trabelsi_deep_2017} we identify $\cplx$ with $\real^2$ via $
  z_\mathrm{r} + j z_\mathrm{i} \mapsto (z_\mathrm{r}, z_\mathrm{i})
$, storing the real and imaginary parts in interleaved format, and emulate $\cplx$-algebra
using $\real$-valued arithmetic. Thus designed $\cplx$-network is, essentially, a
$\real$-network with \textit{double} the layers' widths to accommodate real and imaginary
parts and with constraints on layers' parameters that make them respect $\cplx$-algebra.
This makes it possible to retrofit $\cplx$-valued networks into the existing $\real$-valued
auto-differentiation libraries.

@@

A characterization of a $\cplx$-Gaussian vector, more useful for the current study, is via the
equivalence with the paired real multivariate Gaussian distribution:
\begin{equation}  \label{eq:cn-paired-real-density}
  z \sim \mathcal{N}^{\cplx}_m(\mu, \Gamma, C)
    \Leftrightarrow
    \begin{pmatrix}
      \Re z \\ \Im z
    \end{pmatrix}
      \sim \mathcal{N}_{2 m}\biggl(
        \begin{pmatrix}
          \Re \mu \\ \Im \mu
        \end{pmatrix},
        \frac12
        \begin{pmatrix}
          \Re{(\Gamma + C)} & - \Im{(\Gamma - C)} \\
          \Im{(\Gamma + C)} &   \Re{(\Gamma - C)}
        \end{pmatrix}
      \biggr)
    \,.
\end{equation}
It is possible to recover the $\cplx$ covariance and relation matrices from a block covariance
matrix $\Sigma$ of a \textit{non-degenerate} paired real Gaussian vector through
$$
\begin{pmatrix}
  \Sigma_{rr} & \Sigma_{ri} \\ 
  \Sigma_{ir} & \Sigma_{ii} 
\end{pmatrix}
  = \frac12
    \begin{pmatrix}
      \Re{(\Gamma + C)} & - \Im{(\Gamma - C)} \\
      \Im{(\Gamma + C)} &   \Re{(\Gamma - C)}
    \end{pmatrix}
  \, \Leftrightarrow
  \,
  \begin{cases}
    \Gamma
      = \Sigma_{rr} + \Sigma_{ii} + j (\Sigma_{ir} - \Sigma_{ri})
    \,, \\
    C = \Sigma_{rr} - \Sigma_{ii} + j (\Sigma_{ir} + \Sigma_{ri}) 
  \end{cases}
  $$

@@

To develop $\cplx$-variational dropout we assume a fully factorized posterior approximation
distribution on the layers' parameters: $\omega = (W, b)$, where a $\cplx$-valued $W$ collects
all weight matrices of linear layers and kernels of convolutional layers, and $b$ represents
all bias term combined. We use a fully factorized Dirac family on the biases, and mean-field
proper $\cplx$-Gaussian variational approximation for the weights: $
  q(W) = \prod_{lij} q(W^{l}_{ij})
$ with each parameter being an independent proper complex Gaussian
\begin{equation} \label{eq:c-gauss-vi}
  W_{ij}
    \sim \mathcal{N}^{\cplx}(\mu_{ij}, \sigma^2_{ij}, 0)
  \,.
\end{equation}
Thus we would like to sparsify the weights in the network and, since the bias terms constitute
a meagre share of network's parameters and are rarely dropped out or pruned, we use Bayesan
MAP estimates for them.

@@

\bibliographystyle{amsplain}
\bibliography{references}
\nocite{*}

@@

@@

Gist of the LRP (kingmaetal2015)
translating uncertainty (randomness) from parameters to activations
  adds noise -- helps regularization
  emulates one draw per batch element -- decorrelates gradient within the batch
  reduces computations (memory), leverages batched matmuls (or convos)
  decreases variance of the gradient

@@

\section{Complex Variational Dropout} % (fold)
\label{sec:complex_variational_dropout}

@@

% revirew of kingma and lrt
The key contributions of \cite{kingma_variational_2015} are reinterpretation of dropout
as a variational inference method and the analysis of the effects that the local reparameterization
trick has on the efficiency of the gradient estimator in stochastic variational inference,
\cite{citation_needed}.

\cite{kingmaetal2015} argue that Gaussian dropout, being a multiplicative parameter noise,
can be interpreted as variational inference via factorized Gaussian posterior approximation
$q(\omega)$ that optimizes the $\log$-likelihood term from the Evidence Lower Bound. In a fully
Bayesian treatment, the dropout rate $p \in (0, 1)$, which governs the regularization strength,
would appear in the KL-divergence term of ELBO through the corresponding odds $
  \alpha = \tfrac{p}{1-p}
$, which acts a variance scaling factor in the posterior approximation. Having made $\alpha$
a learnable parameter of the variational approximation, \cite{kingmaetal2015} natuallry propose
letting it vary on the per-parameter basis, which allows for even more flexible posterior
approximations. % who proposed sparsification?

@@

% exposition

% plan

@@

% set by step we retell the story of kingmaetal2015

Gaussian variational approximation enabled \cite{kingma_variational_2015} propose a
\textit{local reparameterization trick}, which reduces the variance of the gradients
in SGVB and does not resample $W$ per element of a minibatch. It relies the closure
of Gaussian distribution under linear transformations. Indeed, if $W$ is a Gaussian
matrix with independently distributed $
  W_{ij} \sim \mathcal{N}(\theta_{ij}, \sigma^2_{ij})
$, then we have \important{moved}.

@@

Maths and related results \cite{pav_moments_2015,taubock_complex-valued_2012},
and \cite{karseras_caution:_2014}

% section complex_variational_dropout (end)

>>>>

%% notation
\newcommand{\real}{\mathbb{R}}
\newcommand{\cplx}{\mathbb{C}}
\newcommand{\tr}[1]{\mathop{tr}{#1}}

@@

%% drafting macro
\newcommand{\important}[1]{\textbf{\color{red} #1}}
\newcommand{\attn}[2]{\textbf{\color{red} #2~\textsuperscript{\textit{[#1]}}}}
\newcommand{\verify}[1]{\attn{verify}{#1}}
\newcommand{\rewrite}[1]{\attn{rewrite}{#1}}
\newcommand{\todo}[1]{{\color{blue} [TODO]} \important{#1}}

%% red-highlight missing citations
\usepackage{etoolbox}
\makeatletter
\patchcmd{\@citex}{\bfseries ?}{\colorbox{red}{\bfseries ?}}{}{}
\makeatother

@@

% complex nets and related stuff
Maths and related results \cite{pav_moments_2015,taubock_complex-valued_2012}, and
\cite{karseras_caution:_2014}

The use $\cplx$-valued networks in applications is not new: earliest mentions include
\cite{citation_needed}. More recently they were applied to music transcription and speech
recognition \cite{trabelsi_deep_2017,yang_complex_2019}. The key motivation behind adoption
of $\cplx$-networks is the fact that complex numbers are inherent in digital signal processing
\cite{citation_needed}, acoustic and radio signal analysis \cite{citation_needed}, and
quantum systems modelling \cite{citation_needed} etc. All these applications require
compliance with algebraic operations in field of complex numbers.

The paper is structured as follows. In sec.~\ref{sec:bayesian_dropout} we review Bayesian
dropout techniques, and ins sec.~\ref{sec:c_valued_networks} we provide a brief summary of
the inner working of complex-valued networks as functions of complex argument. The main
contribution of this study is presented in sec.~\ref{sec:dropout_for_c_valued_layers},
where we consider different variational approximations and priors, outline the tricks
and derive expressions for the divergence term. In sec.~\ref{sec:experiments} we evaluate
the sparsification rate and compare the resulting performance of various $\cplx$-networks
proposed in prior work, and discuss the outcomes in sec.~\ref{sec:discussion_and_related_research}.

@@

For a particular family of variational posterior approximation, variational inference
can be used as a regularization method \cite{kingma_variational_2015} and as model sparsification
technique \cite{molchanov_variational_2017}. In \cite{kingma_variational_2015} the Binary
dropout and DropConnect \cite{hinton_improving_2012,dropconnect}, and Gaussian dropout
\cite{srivastava_dropout_2014,wang_fast_2013} were reinterpreted as variational inference
methods. In that study, authors argue that dropout, being multiplicative parameter noise,
induces parameter distribution equivalent to a fully factorized variational posterior
approximation of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where each component $q_{\theta}(\omega_j)$ is defined via $\omega_j = \mu_j \xi_j$
with $\xi_j \sim p(\xi_j)$ iid for some $p(\xi)$. In Binary dropout with rate $p \in (0, 1)$
the noise $\xi$ is binary and distributed $
  \mathcal{B}\bigl(
      \{0, \tfrac1{1-p}\}, 1-p
    \bigr)
$, whereas in Gaussian dropout $
  \xi \sim \mathcal{N}(1, \alpha)
$ for $\alpha = \tfrac{p}{1-p}$. In the latter case, the \textit{dropout score} $\alpha$ acts
like a variance scaling factor, making it a natural variational parameter for the posterior
approximation, which is optimized via SGVB \eqref{eq:elbo}. Thus \cite{kingma_variational_2015}
propose Variational Dropout technique for Bayesian model regularization. The flexibility in
specifying dependence of $q_\theta$ on $\alpha$ allows imposing structure on dropout: global
rate for the whole model network, local per each component and even specific to individual
parameter.

@@

Bayesian dropout technique with mean field Gaussian variational approximation is further
developed in \cite{molchanov_variational_2017} and \cite{kharitonov_variational_2018}. The
first paper studies variational dropout from a model sparsification perspective and improves
on the results of \cite{kingma_variational_2015} in two directions: the authors propose a
more accurate approximation to the KL-divergence term in \eqref{eq:elbo}, and introduce
additive noise parameterization for the factors of $q_\theta(\omega)$, $\omega = W$.

Under the original $(\mu, \alpha)$ parameterization the stochastic output of a linear
layer $
  y = W^\top x + b
$ with $
  W \in \mathbb{R}^{n\times m}
$ and independent $
  W_{ij} \sim \mathcal{N}(\mu_{ij}, \alpha_{ij} \mu_{ij}^2)
$ is equivalent in distribution to $
  y \sim \prod_i \mathcal{N}\Bigl(
      \sum_j \mu_{ij} x_j + b_i,
      \sum_j \alpha_{ij} \mu_{ij}^2 x_j^2
  \Bigr)
$. \cite{molchanov_variational_2017} argue that for the purposes of model sparsification
this parameterization should be abandoned in favour of an equivalent $(\mu, \sigma^2)$ with
the score $\alpha$ calculated using $\tfrac{\sigma^2}{\mu^2}$ when needed. This makes $
  \nabla_\mu \mathcal{L}^{SGVB}
$, which is particularly important for sparsification, independent of the noise injected
by the trick.

\cite{kingma_variational_2015} and \cite{molchanov_variational_2017} consider a factorized
prior $\pi(\omega)$ with improper log-uniform factors $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$. This unravels the divergence term in \eqref{eq:elbo} into a sum of the following terms:
\begin{equation} \label{eq:improper-kl-div-real}
  KL(q_\theta(\omega) \| \pi(\omega))
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
The expectation in \eqref{eq:expect-improper-term-cplx} is half the $\log$-moment of a non-central
$\chi^2$ distribution with shape $1$ and non-centrality $\lambda = \tfrac1{\alpha}$. In general,
there is no analytically tractable expression for log-moment of a $\chi^2_m$ with odd degrees
of freedom. To circumvent this, \cite{kingma_variational_2015} approximate the expectation by
a non-linear polynomial regression for truncated $\alpha \in (0, 1)$. This approximation was
substantially refined in \cite{molchanov_variational_2017}, both in term of the domain of scores
$\alpha$ and accuracy. The paper considered regression with sigmoid and soft-plus basis functions
fit to the Monte-Carlo estimator of \eqref{eq:improper-kl-div-real} for $\alpha$ varying over
a fine logarithmic grid. In sec.~\ref{sub:real-chisq-grad} we verify the approximation and
compare the approximate derivative to the exact derivative of \eqref{eq:improper-kl-div-real}
with respect to $\log \alpha$.
% Since gradient based methods rely on unbiased gradient estimates

@@

@@

Following \cite{trabelsi_deep_2017}, we naturally identify $\cplx$ with $\real^2$ via $
  z_\mathrm{r} + j z_\mathrm{i} \mapsto (z_\mathrm{r}, z_\mathrm{i})
$, storing the real and imaginary parts in interleaved format, and emulate $\cplx$-algebra
using $\real$-valued arithmetic. Thus designed $\cplx$-network is, essentially, a
$\real$-network with \textit{double} the layers' widths to accommodate real and imaginary
parts and with constraints on layers' parameters that make them respect $\cplx$-algebra.
This makes it possible to retrofit $\cplx$-valued networks into the existing $\real$-valued
auto-differentiation libraries.

@@

A characterization of a $\cplx$-Gaussian vector, more useful for the current study, is via the
equivalence with the paired real multivariate Gaussian distribution:
\begin{multline}  \label{eq:cn-paired-real-density}
  z \sim \mathcal{N}^{\cplx}_m(\mu, \Gamma, C)
    \\\Leftrightarrow
    \begin{pmatrix}
      \Re z \\ \Im z
    \end{pmatrix}
      \sim \mathcal{N}_{2 m}\biggl(
        \begin{pmatrix}
          \Re \mu \\ \Im \mu
        \end{pmatrix},
        \frac12
        \begin{pmatrix}
          \Re{(\Gamma + C)} & - \Im{(\Gamma - C)} \\
          \Im{(\Gamma + C)} &   \Re{(\Gamma - C)}
        \end{pmatrix}
      \biggr)
    \,.
\end{multline}
It is possible to recover the $\cplx$ covariance and relation matrices from a block covariance
matrix $\Sigma$ of a \textit{non-degenerate} paired real Gaussian vector through
$$
\begin{pmatrix}
  \Sigma_{rr} & \Sigma_{ri} \\ 
  \Sigma_{ir} & \Sigma_{ii} 
\end{pmatrix}
  = \frac12
    \begin{pmatrix}
      \Re{(\Gamma + C)} & - \Im{(\Gamma - C)} \\
      \Im{(\Gamma + C)} &   \Re{(\Gamma - C)}
    \end{pmatrix}
  \, \Leftrightarrow
  \,
  \begin{cases}
    \Gamma
      = \Sigma_{rr} + \Sigma_{ii} + j (\Sigma_{ir} - \Sigma_{ri})
    \,, \\
    C = \Sigma_{rr} - \Sigma_{ii} + j (\Sigma_{ir} + \Sigma_{ri}) 
  \end{cases}
  $$

@@

To develop $\cplx$-variational dropout we assume a fully factorized posterior approximation
distribution on the layers' parameters: $\omega = (W, b)$, where a $\cplx$-valued $W$ collects
all weight matrices of linear layers and kernels of convolutional layers, and $b$ represents
all bias term combined. We use a fully factorized Dirac family on the biases, and mean-field
proper $\cplx$-Gaussian variational approximation for the weights: $
  q(W) = \prod_{lij} q(W^{l}_{ij})
$ with each parameter being an independent univariate proper complex Gaussian
% (we omit the subscript $m=1$ for brevity)
\begin{equation} \label{eq:c-gauss-vi}
  W_{ij}
    \sim \mathcal{N}^{\cplx}(\mu_{ij}, \sigma^2_{ij}, 0)
  \,.
\end{equation}
Thus we would like to sparsify the weights in the network and, since the bias terms constitute
a meagre share of network's parameters and are rarely dropped out or pruned, we use Bayesan
MAP estimates for them.

@@

\bibliographystyle{unsrt}
\bibliography{references}
% \nocite{*}

@@

In a fully Bayesian treatment, the dropout rate $p$, which governs the regularization strength,
would appear in the KL-divergence term of ELBO \eqref{eq:elbo} through the corresponding odds 
ratio $
  \alpha = \tfrac{p}{1-p}
$, which.

it vary on the per-parameter basis, which allows for even more flexible posterior
approximations.

% revirew of kingma and lrt
The key contributions of \cite{kingma_variational_2015} are reinterpretation of dropout
as a variational inference method

that optimizes the $\log$-likelihood term from the Evidence Lower Bound.

 % who proposed sparsification?

@@

Gist of the LRP (\cite{kingma_variational_2015})
translating uncertainty (randomness) from parameters to activations
  adds noise -- helps regularization
  emulates one draw per batch element -- decorrelates gradient within the batch
  reduces computations (memory), leverages batched matmuls (or convos)
  decreases variance of the gradient

@@

\section{Bayesian methods} % (fold)
\label{sec:bayesian_methods}

@@

@@

@@

% set by step we retell the story of kingma_variational_2015
Gaussian variational approximation enabled \cite{kingma_variational_2015} propose a
\textit{local reparameterization trick}, which reduces the variance of the gradients
in SGVB and does not resample $W$ per element of a minibatch. It relies the closure
of Gaussian distribution under linear transformations. Indeed, if $W$ is a Gaussian
matrix with independently distributed $
  W_{ij} \sim \mathcal{N}(\theta_{ij}, \sigma^2_{ij})
$, then we have \important{moved}.

@@

% section bayesian_methods (end)
